### 2021-07-08 paper/paper_draft.tex

We supplement these data with a variety of other information.  Most importantly, we add facility-level information including fuel source, ownership, nameplate capacity, heat rate, and estimated emissions-intensity based on heat rates and fuel sources. ADD ON-LINE APPENDIX TABLE OF PLANT INFO.

We also use compliance data from the Alberta Specified Gas Emitters Regulations and the Federal Greenhouse Gas Emissions Reporting Program to provide, where available, more precise emissions intensity data for facilities.  Since these data are both reported at the facility level, not by generating unit, we apply these measurements in the same way in our data.  As such, we do not allow emissions intensity to vary with the intensity of use of a particular unit within a facility in a particular hour nor do we account for any ramping rates of units.

We add weather data from Environment Canada weather stations in Edmonton, Fort McMurray, Cold Lake, and Calgary, covering the three major demand centers in Alberta.  We take the average of available measurements in each area in order to maximize the number of hours we can cover with weather data. With this approach, we lose fewer hours from our data set than if we allow unavailable weather data from one station to remove an hour from our sample. There is also no reason to believe that weather data availability is correlated with our variables of interest, and so the lost observations should have no leverage on the results.

Finally, we merge daily natural gas prices for the Nova Inventory Transfer hub.  We access daily spot prices via NRGStream, a commercial data aggregation service.

All data save the natural gas prices which we are not authorized to redistribute are available HERE and can be replicated using code for the paper which is available at GITHUB LINK. For the publicly available code, we substitute Henry Hub gas prices as a placeholder, to which we apply the average discount observed between Henry Hub and NIT to adjust the values so that regression results may be more closely replicated.

@@

Using our dataset, we can provide a statistical portrait of Alberta's power market. Alberta's wholesale market is a single price, energy-only market. There is no day-ahead energy market, but there is a separate ancillary services market which we do not address beyond a couple of tangential references in this paper.  Alberta's record internal load, reached on February 9, 2021, is 11,729MW, so Alberta's is a relatively small power market compared to the similarly-structure Energy Reliability Council of Texas (ERCOT) market, for example, which had a 58,597 MW peak 2021 load.\footnote{The AESO previous record, reached on January 11, 2018, was 11,697 MW, so peak loads have been quite stable in recent years after a period of rapid growth.} Alberta has seen significant growth in average and peak loads, as shown in Figure \ref{fig:ab_load}, with X cumulative annual average load growth rate during our sample period, although growth has been substantially slower since 2014-2015 due to the effects of depressed oil prices on the Alberta economy discussed further below. There was also a major short-term market event with a wildfire in Fort McMurray in 2016 which took a lot of oil sands generation and load off-line, which explains the markedly-lower spring trough in that year shown in Figure \ref{fig:ab_load}. The 2020 COVID-19 demand shock is also evident in this graph, but lies outside the sample period for our analysis.

@@

Our sample period covers ten years with no major regulatory changes in the power market design. Alberta briefly considered an addition of a capacity market, but this was never implemented. However, market conditions have varied over the sample period. There are three distinct periods during our sample - relatively high load growth and tight market conditions from 2009 through 2013, followed by sharply constrained growth and high reserve margins after 2014 through most of 2018, followed by a return to tighter market conditions after the spring of 2018. These conditions are reflected in peak and off-peak prices shown in Figure \ref{fig:ab_prices}. Some of these market changes coincide with policy changes in our experiment - in particular, the changes to the \emph{SGER} introduced in June of 2015 which took effect in January of 2016 and 2017 respectively coincide with a period of over-supply, while the introduction of the \emph{CCIR} in 2018 occurred during a period of tightening reserve margins and lingering uncertainty over market structure. From 2018 through pre-COVID 2020, both GHG policy and the wholesale electricity market were tightening, and so we need to be careful to disentangle these impacts.

\begin{figure}[t]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../images/peak_prices.png}
\vspace{-0.75cm}	\caption{Wholesale power prices, peak and off-peak hours. Peak hours are 8am to 11pm other than on statutory holidays or Sundays. Source: Alberta Electric System Operator data.}
\label{fig:ab_prices}
\end{figure}

\begin{figure}[t]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../images/gen_fuel.png}
\vspace{-0.75cm}	\caption{Generation mix in Alberta.  Source: Alberta Electric System Operator data.}
\label{fig:gen_mix}
\end{figure}

The generation mix in Alberta's power market is dominated by fossil fuels although, as shown in Figure \ref{fig:gen_mix}, the dominant fossil fuel has changed from coal to natural gas. Within natural gas generation, the mix of plant types is also relevant to our study. By the end of our sample period, X\% of installed natural gas capacity comes from combined cycle (NGCC) plants, while y\% is from simple-cycle (NGSC) plants, and the remaining Z\% of capacity is from combined heat and power (COGEN) facilities. Combined heat and power plants tend to be price-takers in the market, with very little flexibility at the margin since the industrial processes with which they are associated rely on them for process heat. As will be discussed below, net-to-grid power from these facilities is primarily offered into the market at a \$0 offer and accepts market price, and so we see limited evidence of carbon price pass-through for most of the power offered by these facilities.

ADJUST FIGURE TO SHOW SAMPLE PERIOD

@@

The nature of the wholesale market is such that different generators will derive different levels of average revenue from the sale of their power. Those that dispatch only in high-price times will have higher revenues on average than plants without dispatch control, or plants which run most hours of the day. Since prices have varied substantially over time, as shown above in Figure \ref{fig:ab_prices}, captured average prices at the plant level have varied as well, as shown in Figure \ref{fig:capture_prices}. We can see evidence here of how plants of different types offer their power into the market: hydro and combined- and simple-cycle gas turbines follow prices, and dispatch more capacity in higher-priced hours, just as imports tend to occur when prices are higher. Lower prices tend to prevail when wind generation is highest, and so captured prices for wind power facilities are lower than other forms of generation, and exports tend to be scheduled when prices are expected to be low. Solar, a relatively new entrant in the Alberta market, has captured higher prices than other variable renewable sources since its generation pattern coincides with daily and seasonal peak prices, although we would expect this to change as the \emph{duck curve} effects take hold over time.

\begin{figure}[!h]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../images/price_capture.png}
\vspace{-0.75cm}	\caption{Annual average revenue by plant type}
\label{fig:capture_prices}
\end{figure}

@@

Alberta's three carbon pricing regimes affect generating facilities in two ways. First, each of three (\emph{SGER}, \emph{CCIR}, and \emph{TIER}) regimes imposes a price on emissions. Second, each regime allocates emissions credits based on output or deemed avoided emissions. The net effect of these policies creates variation within and across facility types over time.  Consider, for example, Figure \ref{fig:sector_compliance_cost} which shows the mean value and range of annual compliance costs for larger generators in Alberta's fleet over the sample period.\footnote{Plants were only included in the sample for this figure if their annual emissions were sufficiently large ($\geq$100kt CO$_2$e per year) such that we have specific compliance cost data. For our estimate, we impute compliance costs for a wider set of facilities using emissions intensities from the compliance cost data.}

\begin{figure}[!h]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../../SGER/sector_compliance_cost.png}
\vspace{-0.75cm}	\caption{Annual average revenue by plant type}
\label{fig:sector_compliance_cost}
\end{figure}

As Figure \ref{fig:sector_compliance_cost} shows, changes in facility compliance costs were not all coincident with increased marginal carbon prices, but are actually dominated by the changes in output-based allocations which occurred in 2018-2019 rather than the initial increase carbon prices from \$15 to \$30 per tonne \coe between 2016 and 2018.

\section{Estimation Strategy}
There are two factors which confound our ability to draw inference from the merit order data. The first is that plants of the same type (e.g. coal or combined-cycle natural gas plants) may be of different sizes (e.g. a 350MW plant vs a 200MW plant) while facing similar optimization constraints, and they may be jointly dispatched by individual operators. The second issue is that the strategy space for each facility is very large, given that they may decide both on the break-points for each of up to 7 blocks of power, and then decide on a price for the first block and an adder for each subsequent block offered into the market. For each hour, then, the optimization problem consists of 12 choice variables and we have data on these choices for over 18.5 million facility-hour-block pairs. We render our problem more parsimonious, and at the same time overcome the portfolio bidding issue, both how the overall merit order and how power from a particular type of plant has been offered into the market over time.

This is best understood with a series of figures, and for this we will again use one coal plant (Capital Power's Genessee 3 facility) as well as Alberta's overall coal fleet to demonstrate the approach we take here.

@@

\section{Analysis and Results}

>>>>

We supplement these data with a variety of other information.  Most importantly, we add facility-level compliance data from Alberta's greenhouse gas emissions pricing policies, the SGER (2009-2017) and the CCIR (2018-2019). These data were provided for this analysis by the Government of Alberta, with some portions of the data also available publicly. We supplement these data with information from the Federal Greenhouse Gas Emissions Reporting Program to provide, where available, more precise emissions intensity data for smaller facilities.\footnote{The federal data are useful since the reporting threshold of 10kt \coe per annum is lower than the provincial threshold for coverage under the emissions pricing policies of 100kt\coe per annum.}  Since these data are both reported at the facility level, not by generating unit, we apply these measurements in the same way in our paper and so we cannot differentiate between multiple generating units on particular sites.\footnote{There are two exceptions to this. The Keephills 3 and Genessee 3 coal-fired generating units report separately under Alberta's carbon pricing rules, and so we have separate emissions-intensity data for them.}  Since we have only annual data, we do not allow emissions intensity to vary with the intensity of use of a particular unit within a facility, nor do we allow emissions intensities to vary with ramping of units.

We add weather data from Environment Canada weather stations in Edmonton, Fort McMurray, Cold Lake, and Calgary, covering the major demand centers in Alberta.  We take the average of available measurements in each area in order to maximize the number of hours we can cover with weather data. With this approach, we lose fewer hours from our data set than if we allow unavailable weather data from one station to remove an hour from our sample. There is also no reason to believe that weather data availability is correlated with our variables of interest, and so the lost observations should have no leverage on the results. We convert temperature data to heating and cooling degree days for each region, which we then use in our model.

Finally, we supplement our publicly-available data with several series from NRGStream, a commercial data aggregation service. Most importantly, this service provides daily natural gas prices for the Nova Inventory Transfer hub which are a better proxy for Alberta natural gas prices than publicly-available series for Henry Hub. NRGStream also scrapes the real-time data from the AESO, and thus provides slightly different data from the aggregate, publicly-available historic metered volumes provided by the grid operator. We use NRGStream real-time data to compile real-time intertie flows between Alberta and British Columbia, Saskatchewan, and Montana and to source real-time generation (as opposed to net-to-grid metered volumes) from all generating units in the province where needed.

All data save the natural gas prices and NRGStream generation and trade date which we are not authorized to redistribute are available HERE and can be replicated using code for the paper which is available at GITHUB LINK. For the publicly available code, we substitute Henry Hub gas prices as a placeholder, to which we apply the average discount observed between Henry Hub and NIT to adjust the values so that regression results may be more closely replicated.

@@

A statistical portrait of Alberta's power market provides context for our study. Alberta's wholesale market is a single price, energy-only market. There is no day-ahead energy market, but there is a separate ancillary services market which we do not address beyond a couple of tangential references in this paper.  Alberta's record internal load, reached on February 9, 2021, is 11,729MW, so Alberta's is a relatively small power market compared to the similarly-structured Energy Reliability Council of Texas (ERCOT) market, for example, which had a 58,597 MW peak 2021 load. Alberta has seen significant growth in average and peak loads, as shown in Figure \ref{fig:ab_load}, although growth has been substantially slower since 2014-2015 due to the effects of depressed oil prices on the Alberta economy discussed further below. There was also a major, short-term market event with a wildfire in Fort McMurray in 2016 which took a lot of oil sands generation and load off-line, which explains the markedly-lower spring trough in that year's load profile as shown in Figure \ref{fig:ab_load}. The 2020 COVID-19 demand shock is also evident in Figure \ref{fig:ab_load}, but lies outside the sample period for our analysis.

@@

Our sample period covers ten years with no major regulatory changes in the power market design. Alberta briefly considered an addition of a capacity market, but this was never implemented. However, market conditions have varied over the sample period. There are three distinct periods during our sample - relatively high load growth and tight market conditions from 2009 through 2013, followed by sharply constrained growth and high reserve margins after 2014 through most of 2018, followed by a return to tighter market conditions after the spring of 2018. These conditions are reflected in peak and off-peak prices shown in Figure \ref{fig:ab_prices}. Some of these market changes coincide with policy changes of interest in our paper, in particular the changes to the \emph{SGER} introduced in June of 2015 which took effect in January of 2016 and 2017 respectively and are coincidental with a period of over-supply, and the introduction of the \emph{CCIR} in 2018 which occurred during a period of tightening reserve margins and lingering uncertainty over market structure. From 2018 through pre-COVID 2020, both GHG policy and the wholesale electricity market were tightening, and so we need to be careful to disentangle these impacts.

\begin{figure}[t]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../images/peak_prices.png}
\vspace{-0.75cm}	\caption{Wholesale power prices, peak and off-peak hours. Peak hours are 8am to 11pm other than on statutory holidays or Sundays. Source: AESO data, authors' graph.}
\label{fig:ab_prices}
\end{figure}

\begin{figure}[t]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../images/gen_area_grey.png}
\vspace{-0.75cm}	\caption{Generation mix in Alberta.  Source: AESO data provided by NRGStream, authors' graph.}
\label{fig:gen_mix}
\end{figure}

The generation mix in Alberta's power market is dominated by fossil fuels although, as shown in Figure \ref{fig:gen_mix}, the dominant fossil fuel has changed from coal to natural gas. Within natural gas generation, the mix of plant types is also relevant to our study. By the end of our sample period, a significant increase in generating capacity from combined cycle (NGCC) plants had been installed, and this is reflected in higher generation from those sources. The installed capacity of combined heat and power (COGEN) facilities also grew during the sample period, and these plants tend to operate with high capacity factors but also tend to be price-takers in the market, with very little flexibility at the margin since the industrial processes with which they are associated rely on them for process heat. As will be discussed below, net-to-grid power from COGEN facilities is primarily offered into the market at a \$0 offer and accepts market price, and so we see limited evidence of carbon price pass-through for most of the power offered by these facilities. On the other hand, peaking capacity in the market during our sample period was largely met through ramping of coal and combined-cycle plants, as well as simple cycle (SCGT) natural gas turbines. Each of these would be exposed to carbon pricing, and so we expect to see pass through affect offers of power from these facilities into the market.

@@

The nature of Alberta's wholesale market is such that different generators will derive different levels of average revenue from the sale of their power. Those that dispatch only in high-price times will have higher revenues on average than plants without dispatch control, or plants which run most hours of the day. Since prices have varied substantially over time, as shown above in Figure \ref{fig:ab_prices}, captured average prices at the plant level have varied as well, as shown in Figure \ref{fig:capture_prices}. We can see evidence here of how plants of different types offer their power into the market: hydro and combined- and simple-cycle gas turbines follow prices, and dispatch more capacity in higher-priced hours, just as imports tend to occur when prices are higher. Lower prices tend to prevail when wind generation is highest, and so captured prices for wind power facilities are lower than other forms of generation, and exports tend to be scheduled when prices are expected to be low. Solar, a relatively new entrant in the Alberta market, has captured higher prices than other variable renewable sources since its generation pattern coincides with daily and seasonal peak prices, although we would expect this to change as \emph{duck curve} effects take hold over time.

\begin{figure}[t]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../images/price_capture.png}
\vspace{-0.75cm}	\caption{Annual average revenue by plant type. source: AESO data provided by NRGStream, authors' graph.}
\label{fig:capture_prices}
\end{figure}

@@

Alberta's three carbon pricing regimes each affected generating facilities in two ways. First, each of three (\emph{SGER}, \emph{CCIR}, and \emph{TIER}) regimes imposes a price on emissions. Second, each regime allocates emissions credits based on output or deemed avoided emissions. The net effect of these policies creates variation within and across facilities and facility types and over time.  Consider, for example, Figure \ref{fig:sector_compliance_cost} which shows the mean value and range of annual compliance costs for larger generators in Alberta's fleet over the sample period.\footnote{Plants were only included in the sample for this figure if their annual emissions were sufficiently large ($\geq$100kt CO$_2$e per year) such that we have specific compliance cost data. For our estimation results, we impute compliance costs for a wider set of facilities using emissions intensities from the compliance cost data.}

\begin{figure}[t]%%
	\centering \vspace{-.25cm} \includegraphics[width=6.5in]{../../SGER/sector_compliance_cost.png}
\vspace{-0.75cm}	\caption{Upper: Annual average compliance cost (line) and range (shaded area) by plant type. Lower: Price charged on excess net annual emissions under Alberta's carbon pricing programs. Source: Alberta Environment compliance cost data}
\label{fig:sector_compliance_cost}
\end{figure}

As Figure \ref{fig:sector_compliance_cost} shows, changes in facility compliance costs were not all coincident with increased marginal carbon prices, but are actually dominated by the changes in output-based allocations which occurred in 2018-2019 rather than the initial increase carbon prices from \$15 to \$30 per tonne \coe between 2016 and 2018. The largest cost increases affect coal-fired power plants and some simple-cycle generators, while have limited effects on combined heat and power plants. Compliance costs for natural gas combined-cycle plants dropped between 2017 and 2019 as a result of a change in the formula by which output-based allocations were calculated.

\section{Estimation Strategy}
There are two factors which confound our ability to draw inference from the merit order data. The first is that plants of the same type (e.g. coal or combined-cycle natural gas plants) may be of different sizes (e.g. a 350MW plant vs a 200MW plant) while facing similar optimization constraints, and they may be jointly dispatched by the same operator. In particular for coal plants, plant-level minimum-must-run constraints imply that facilities will always offer some portion of their generation into the market at a \$0/MWh price, effectively acting as price takers for that hour for that block of power. They will offer marginal blocks of power, for capacity above their minimum-must-run level, at higher prices. However, given that plants are of different sizes, and blocks are endogenously determined, it's challenging to identify a change in offers looking only at block offers. The second issue is that the strategy space for each facility is very large, given that they may decide both on the break-points for each of up to 7 blocks of power, and then decide on a price for the first block and an adder for each subsequent block offered into the market. For each hour, then, the optimization problem consists of 12 choice variables and we have data on these choices for almost 20 million facility-hour-block pairs. Finally, since multiple plants may be controlled by a single entity, the observed data are the product of a complex portfolio optimization problem and the offers of any individual plant may not reveal an overall portfolio-level attempt to pass through the costs of carbon prices to wholesale prices. We avoid these problems by constructing synthetic offer curves either by plant-type or by controlling entity, thus estimating the impact of policy changes on the supply behaviour of the aggregate fleet rather than on any individual plant. We can then use the synthetic merit order data we construct to examine how power has been offered into the system over time as climate policies have varied.

Our empirical strategy is best understood visually through a series of figures. Consider Figure \ref{fig:synth_merit}. In the left-hand panel, we show the merit order for a particular hour for all of the coal facilities in the province, with offers from multi-unit facilities combined for ease of visualization. We see facilities bidding some of their power at \$0, ensuring they are in the market, with marginal generation blocks offered at increasing marginal prices. In this particular hour, the price was such that most of the offered coal generation was dispatched, with only blocks offered above the \$802/MWh market clearing price not being dispatched. In the right hand panel, we show our synthetic merit order which summarizes all of the coal power offered into the market in a given hour using the 15 points shown in the Figure, which correspond to the value of the merit order at prescribed percentiles of the total offered power in that hour.\footnote{We store the 10th through 40th percentiles in 10\% intervals and the 50th percentile and above in 5\% intervals to better capture the curvature of the merit order.} Summarizing the data this way reduces our stored information by a factor of three, but more importantly it allows us to answer the relevant question in our study - by how much did the supply curve shift in response to changes in the value of carbon pricing costs and output-based allocation revenue - by looking at changes in the value of our sampling points across the portfolio of offered power. Or, visually, we want to ask how, on average, each of the points in the right-hand panel of Figure \ref{fig:synth_merit} change when carbon prices are increased or when output-based allocations of emissions credits change.

\begin{figure}[!htb]
    \centering
    \begin{minipage}{.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{../images/coal_merit.png}
        %\caption{$dt=0.1$}
        %\label{fig:prob1_6_2}
    \end{minipage}%
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{../images/coal_synth_merit}
        %\caption{$dt =$}
        %\label{fig:prob1_6_1}
    \end{minipage}
    \caption{Observed coal facility merit order (left panel) and synthetic coal merit order (right panel) for February 4, 2019 at 7pm. For each hour, we convert the empirical merit order to a step function, and then extract the y-intercept as at 14 unique percentiles of total offered power in that hour, using 5\% increments in the upper half of the merit order.}
    \label{fig:synth_merit}
\end{figure}

There is a lot of day-to-day and hour-to-hour variability in how coal are offered into the market. Figure \ref{fig:synth_coal_2019}, which shows all of the 7pm merit orders (left panel), their synthetic equivalents, and the average synthetic merit order for the last year of our sample, provides a sense of the degree of variability in our sample. Similar variability exists in the offers of generation from other dispatchable sources.\footnote{See online graphical appendix, Figure A1, for a similar Figure for all dispatchable generating source types.} Figure \ref{fig:synth_coal_2019} shows the degree to which changes in total offered capacity would skew our results if we attempted to base our analysis on offered megawatts. Instead, ny using percentiles of offered capacity as our units of measure, we can more cleanly ask how different segments of the supply curve shift vertically in response to different events, while implicitly assuming that the total offered power available in the market in a given hour is independent of the carbon pricing policy on offer at the time.

\begin{figure}[!htb]
    \centering
    \begin{minipage}{.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{../images/all_coal_merit.png}
        %\caption{$dt=0.1$}
        %\label{fig:prob1_6_2}
    \end{minipage}%
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{../images/coal_synth_2019}
        %\caption{$dt =$}
        %\label{fig:prob1_6_1}
    \end{minipage}
    \caption{Observed coal facility merit order (left panel) and synthetic coal merit order (right panel) for all of the 7pm hours of 2019.}
    \label{fig:synth_coal_2019}
\end{figure}

The next stage in our analysis is to transfer our information on facility compliance costs to the synthetic merit orders. Recall that, for the major emitters, we have specific facility-level information on their emissions-intensity their output-based allocations of emissions credits which determines their net emissions pricing liabilities. Since we are interested in the separate effects of both components of emissions policies, and since both vary at the facility level through our sample, we can separately identify the impacts of each on price pass-through behaviour. As such, for each of the synthetic merit orders we create for analysis purposes, we also capture and store both the marginal emissions charge and the marginal output-based allocation at each percentile support point.

For our analysis sample, we create three other types of synthetic merit orders. First, for very general results, we characterize the entire merit order using the methodology outlined above. We also use the fact that, from 2013 through 2019, the AESO provides offer control by unit as part of their merit order releases. We use this information to construct synthetic supply curves for each of the major participants in the Alberta power market. This means that we have facility-level, generator-class-level and market-level synthetic merit orders to analyze.

To these, synthetic merit orders, we add our hourly-level data on market status, weather, and fuel prices specified above. This then allows us to build regression equations for each percentile level of each synthetic merit order, asking the degree to which carbon price and output-based allocation changes impact the level of each step of the supply curves by facility, plant-type, by owner, or for the complete market.

@@

\subsection{Facility-level impacts}

\subsection{Impacts By Generator Type}

\subsection{Impacts By Generator Type}

\subsection{Market-wide impacts}
